Comparing Results with Mafessoni et al.

### 2019-07-08 LatexManuscript/legacy.tex
pushing changes

\todo{A recent publication by Mafessoni et al. also identified a batch effect in the 1kGP using a method that is independent of quality metrics \citep{mafessoni2018turning}.
They identify over 15,000 variants significantly associated to abnormal LD patterns that are not associated to $Q$ 
\sgcomment{the previous phrase is confusing to me. E.g., does "that" stand for patterns or variants?
Maybe replace by :  
A recent publication by Mafessoni et al. also identified a batch effect in the 1kGP using a method that uses linkage disequilibrium rather than quality metrics to identify over 15,000 suspicious variants \citep{mafessoni2018turning}.
  } (Figure \ref{Mafessoni_MutSpect}).
We find that 3\% of the variants they identified are also in our list of suspicious variants and have correlated p-values to those identified using the GCAT method \ref{Mafessoni_MutSpect}.
We also find 23,172 $Q$-associated variants that are not associated to abnormal LD patterns. \sgcomment{Present proportions for both, or counts for both? Did they use only strict mask? }
Interestingly, the variants identified my Mafessoni et al. are not enriched in the mutational spectra described above (Figure \ref{Mafessoni_MutSpect}).
These results indicate that there may a multitude of batch effects in the 1kGP that can only be identified using a suite of association tests.}

@@

We searched the literature for any GWAS that might have reported these $Q$-associated variants as being significantly associated with some biological trait, even though there is no particular reason for these variants to be associated with phenotypes.
The NHGRI-EBI Catalog of published genome-wide association studies identified seventeen recent publications that had reported these variants as close to or above the genome-wide significant threshold (Table \ref{gwasTable}).
Some of these publications noted that these variants were not flanked by LD peaks \citep{Mandage2017}.
Indeed, we expect a large number of these variants to be filtered out by some quality controls like the Hardy-Weinberg  equilibrium test.
However, the fact that some of these variants are not removed and that the great majority of these variants are missing from higher quality datasets means that these $Q$-associated variants should be flagged for removal to avoid spurious association.

@@

These articles used a variety of strict quality filters, including Hardy-Weinberg equilibrium test, deviations in expected allele frequency and sequencing data quality thresholds.
They also removed rare alleles and alleles with high degrees of missingness.
Even though the studies used state-of-the-art quality controls, the variants were imputed onto genotype data and reached genome wide significance for association with biological traits.

These associations are not necessarily incorrect -- a weak but significant bias in imputation may still result in a correct associations.
To distinguish between variants with weak but significant association with $Q$ from variants with strong biases, we distinguished between variants where the allele frequency difference between individuals with low- and high-$Q$ is larger than a factor of two (which naturally separates two clusters of variants on Figure \ref{90HanSFS_full}).
The majority (92.7\%) of the $Q$-associated variants are strongly biased in that they are more than twice as frequent in individuals with low-$Q$ compared to high-$Q$ data.
By contrast, most $Q$-associated variants reported in the GWAS catalogue had weak bias (See Figure \ref{Pub_30}), with three exceptions.
One study reports associations with seven $Q$-associated variants that we find to be highly biased \citep{Mandage2017}. 
That study considered copy number of Epstein-Barr virus (EBV) sequence in the 1kGP as a phenotype. 
\todo{It is plausible that the EBV copy number phenotype is sensitive to mapping and in this study, may be confounded by the same technical artefacts that lead to biased SNP calling.}

@@

Recommendations : doing their approach as well as ours. We recommend removing the variants we ID and the ones they find. Regression with QC metrics (shoutout to them)

@@

For analyses where individual variants cannot be examined individually (mutation profiles, distributions of allele frequencies, polygenic risk scores), we would recommend to simply discard the $Q$-associated SNPs or the individuals with $Q<30$ (lists of such variants and sample IDs are provided in the Supplementary Data). We also recommend that imputation servers discard individuals with low Q (or at least provide the option of performing the imputation without). Given the value of freely accessible data, resequencing individuals with low Q would  also likely be a worthwhile investment for the community.

@@

On a technical front, we were surprised that strong association between variants and technical covariates in the 1kGP project had not been identified before. 
The genome-wide logistic regression analysis of genotype on quality metric is straightforward, and should probably be a standard in a variety of -omics studies. The logistic factor analysis is more computationally demanding but produces more robust results \citep{song2015testing}. Both approaches produce comparable results.

@@

@@

\begin{figure}[tbp]
\centering
\includegraphics[width=13cm,keepaspectratio]{../Figures/Mafessoni_MutSpect.jpg}
\caption{ \textbf{A} Comparison of the p-values from the association to unusual LD from \cite{mafessoni2018turning} and the p-values from the GCAT association test to quality metrics.
There are a total of 40,266 variants in this plot with only 1328 ( 3\% ) are statistically significant in both test.
There are 23172  variants ( 58\% ) that are $Q$-associated but not LD-associated and 15766 ( 39\% ) that are LD-associated but not $Q$-associated.
\textbf{B} Mutation spectrum of the variants significantly associated to unusual LD from \cite{mafessoni2018turning}. }  
\label{Mafessoni_MutSpect}
\end{figure}

>>>>

A recent publication by Mafessoni et al. also identified a batch effect in the 1kGP using a method that uses linkage disequilibrium rather than quality metrics to identify nearly 20,000 suspicious variants \citep{mafessoni2018turning} (Figure \ref{Mafessoni_MutSpect}).
They identify 17,917 variants significantly associated to abnormal LD patterns that are not associated to $Q$ \ref{Mafessoni_MutSpect}.
We find that 3\% of the variants they identified are also in our list of suspicious variants and have correlated p-values to those identified using the GCAT method \ref{Mafessoni_MutSpect}.
We also find 23,172 $Q$-associated variants that are not associated to abnormal LD patterns. \sgcomment{Present proportions for both, or counts for both? Did they use only strict mask? }
Interestingly, the variants identified my Mafessoni et al. are not enriched in the mutational spectra described above (Figure \ref{Mafessoni_MutSpect}).
These results indicate that there may a multitude of batch effects in the 1kGP that can only be identified using a suite of association tests.

@@

We searched the literature for any GWAS that might have reported these $Q$-associated variants as being significantly associated with some biological trait, even though there is no particular reason for these variants to be associated with phenotypes.
The NHGRI-EBI Catalog of published genome-wide association studies identified seventeen recent publications that had reported these variants as close to or above the genome-wide significant threshold (Table \ref{gwasTable}).

@@

\todo{All of these articles used a variety of strict quality filters, including Hardy-Weinberg equilibrium test, deviations in expected allele frequency and sequencing data quality thresholds.
They also removed rare alleles and alleles with high degrees of missingness.
Indeed, we expect a large number of $Q$-associated variants to be filtered out by some quality controls like the Hardy-Weinberg equilibrium test.
Even though the studies used state-of-the-art quality controls, the variants were imputed onto genotype data and reached genome wide significance for association with biological traits.
However, the fact that some of these variants in other studies are not removed and that the great majority of these variants are missing from higher quality datasets means that these $Q$-associated variants should be flagged for removal to avoid spurious association.

These associations are not necessarily incorrect -- a weak but significant bias in imputation may still result in a correct associations.
To distinguish between variants with weak but significant association with $Q$ from variants with strong biases, we distinguished between variants where the allele frequency difference between individuals with low- and high-$Q$ is larger than a factor of two (which naturally separates two clusters of variants on Figure \ref{90HanSFS_full}).
The majority (92.7\%) of the $Q$-associated variants are strongly biased in that they are more than twice as frequent in individuals with low-$Q$ compared to high-$Q$ data.
By contrast, most $Q$-associated variants reported in the GWAS catalogue had weak bias (See Figure \ref{Pub_30}), with three exceptions.
One study that reports associations with seven highly biased $Q$-associated variants considered copy number of Epstein-Barr virus (EBV) sequence in the 1kGP as a phenotype \citep{Mandage2017}. 
The authors noted that these seven variants were not flanked by LD peaks and were correctly removed from downstream analyses -- though still included in the NHGRI-EBI Catalog \citep{Mandage2017}.
It is plausible that the EBV copy number phenotype is sensitive to mapping and confounded by the same technical artefacts that lead to biased SNP calling.}

@@

@@

For analyses where individual variants cannot be examined individually (mutation profiles, distributions of allele frequencies, polygenic risk scores), we would recommend to simply discard the individuals with $Q<30$ and to filter out the LD-associated variants identified by \cite{mafessoni2018turning} and the $Q$-associated variants we identified (lists of such variants and sample IDs are provided in the Supplementary Data). We also recommend that imputation servers discard individuals with low Q (or at least provide the option of performing the imputation without). Given the value of freely accessible data, resequencing individuals with low Q would  also likely be a worthwhile investment for the community.

@@

On a technical front, we were surprised that strong association between variants and technical covariates in the 1kGP project had not been identified before. 
The genome-wide logistic regression analysis of genotype on quality metric is straightforward, and should probably be a standard in a variety of -omics studies. 
The logistic factor analysis is more computationally demanding but produces more robust results \citep{song2015testing}. 
All three approaches accounting for population structure we tested produce comparable results.

@@

\section{Author's Note}
\todo{While reviewing our manuscript for publication, in response to our pre-print, the New York Genome Center announced that they had resequenced the original 1kGP dataset in a yet to be published dataset.}

@@

\begin{figure}[tbp]
\centering
\includegraphics[width=13cm,keepaspectratio]{../Figures/Mafessoni_MutSpect.jpg}
\caption{ \textbf{A} Comparison of the p-values from the association to unusual LD from \cite{mafessoni2018turning} and the p-values from the GCAT association test to quality metrics.
There are a total of 42,307 variants in this plot with only 1,279 ( 3\% ) are statistically significant in both test.
There are 23,111  variants ( 55\% ) that are $Q$-associated but not LD-associated and 17,917 ( 42\% ) that are LD-associated but not $Q$-associated.
\textbf{B} Mutation spectrum of the variants significantly associated to unusual LD from \cite{mafessoni2018turning}. }  
\label{Mafessoni_MutSpect}
\end{figure}

### 2019-07-08 PlottingScripts/ProcessedData/ScoopCheck.Rmd
pushing changes

load data
```{r}
#load regression results
Regression <- readRDS('/Users/luke/Documents/Legacy_Misc/Misc/Regression.Rmd')
#load supplementary files from Mafessoni
t4 <- fread('/Users/luke/Downloads/evy199_supp/TableS4.bed',skip=1)

#t5 <- fread('/Users/luke/Downloads/evy199_supp/TableS5.bed',skip=1)

t6 <- fread('/Users/luke/Downloads/evy199_supp/TableS6.bed',skip=1)

#t7 <- fread('/Users/luke/Downloads/evy199_supp/TableS7.bed',skip=1)

#t8 <- fread('/Users/luke/Downloads/evy199_supp/TableS8.bed',skip=1)

@@

#totalTus <- merge(Regression, totalT ,by.x = c('Chr','Pos'),by.y=c('V1','V3'))
totalTusX <- merge(Regression, totalT ,by.x = c('Chr','Pos'),by.y=c('V1','V3'), all.x=T)
totalTusX[is.na(totalTusX)] <- 1

totalTusX
```

@@

to_plot <- totalTusX %>%
           filter(V5 < 1 | Adjusted_Log10p > 2)

@@

p_comp <- ggExtra::ggMarginal(p_comp, type = "histogram")

>>>>

load data
```{r}
#load regression results
Regression <- readRDS('/Users/luke/Documents/Legacy_Misc/Misc/Regression.Rmd')
#load supplementary files from Mafessoni
#t5 <- fread('/Users/luke/Downloads/evy199_supp/TableS5.bed',skip=1)
#t7 <- fread('/Users/luke/Downloads/evy199_supp/TableS7.bed',skip=1)
#t8 <- fread('/Users/luke/Downloads/evy199_supp/TableS8.bed',skip=1)

t4 <- fread('/Users/luke/Downloads/evy199_supp/TableS4.bed',skip=1)
t6 <- fread('/Users/luke/Downloads/evy199_supp/TableS6.bed',skip=1)

@@

Merged_Regression <- merge(Regression, totalT ,by.x = c('Chr','Pos'),by.y=c('V1','V3'), all.x=T)
Merged_Regression[is.na(Merged_Regression)] <- 1

Merged_Regression
```

@@

to_plot <- Merged_Regression %>%
           filter(V5 < 1 | Adjusted_Log10p > 2)

@@
